---
title: "tableone: Getting started"
output: html_document
vignette: >
  %\VignetteIndexEntry{tableone: Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

here::i_am("vignettes/tableone.Rmd")
source(here::here("vignettes/vignette-utils.R"))

original_opts = options()
.Last = function() {
  options(original_opts)
}

options(tableone.quiet = TRUE)

library(tidyverse)
library(tableone)
library(survival)
```

# Formula versus `tidyselect` interface

`tableone` is implemented to allow a subset of columns in a large dataset to be
pulled into a table without any fuss. It is also designed with a workflow in
mind that involves building statistical models from the data later. We assume
the data follows a general pattern in that there are one observation per row,
individual columns are specific data points in those observations and may be one
of:

* `outcome`: something that we will be assessing in a statistical model, maybe a 
continuous outcome, or a time measure, or a logical measure.
* `intervention`: the thing that is varied between the different observations
* `covariates`: the other factors that may influence the outcome that we want to 
control for.

In the end we will want to construct a model that takes the following high level
structure:

`outcome ~ intervention + covariate_1 + covariate_2 + ... + covariate_n`

## Simple population description example

Before we build a model we need to firstly compare the distribution of the
covariates in the population and secondly compare them in the intervention and
non-intervention groups, usually done without reference to outcome. To demonstrate
this we are using the `survival::cgd` data set.


```{r}
gcd = survival::cgd %>% 
  # filter to include only the first visit
  filter(enum==1) %>% 
  # make the steroids and propylac columns into a logical value
  # see later for a better way of doing this.
  mutate(
    steroids = as.logical(steroids),
    propylac = as.logical(propylac)
  )
  

# A basic unstratified population description table is as follows:
gcd %>% describe_population(everything())
```

This could have been specified using the formula interface. In this example we
have taken an example of the formula we might wish to use for a survival model
and we reuse it to give us a more targetted descriptive table. It is also 
possible to supply `tableone` with a relabelling function that maps column names
to printable labels, as demonstrated here:

```{r}
# define a formula - this might be reused in model building later
formula = Surv(tstart, tstop, status) ~ treat + 
  sex + age + height + weight + inherit + steroids + hos.cat

# set a table relabelling function
rename_cols = function(col) {
  case_when(
    col == "hos.cat" ~ "Location",
    col == "steroids" ~ "Steroid treatment",
    TRUE ~ stringr::str_to_sentence(col)
  )
}
options("tableone.labeller"=rename_cols)

# create a simple description
gcd %>% describe_population(formula)

```

The relabelling function can either be passed to each invocation of `tableone`
functions or as an option as shown here, which makes the labeller available to
all subsequent calls. This is useful if you are generating many tables from
a single dataset.

We will generally use the formula interface from here on but for exploration
of larger datasets with more covariates the `tidyselect` interface may be
more useful.

## Comparing the population by intervention

In this example a more useful table compares the treatment groups. We can use the
same formula syntax for this, but in this case the first predictor is assumed to
be the intervention and the data set is compared by intervention (in this case
the `treat` column). From this we can conclude that the population is well
distributed between placebo and treatment groups and there is no major bias in 
the randomisation process:

```{r}

# same as above
formula = Surv(tstart, tstop, status) ~ treat + 
  sex + age + height + weight + inherit + steroids + hos.cat

# labelling function is still active
gcd %>% compare_population(formula)

```

Alternatively if we were using the `tidyselect` interface this alternate syntax
would have given us the same table. Note that we must group the data by
intervention, for the `tidyselect` to work as intended:

```R
gcd %>% group_by(treat) %>% 
  compare_population(sex,age,height,weight,inherit,steroids,hos.cat)
```

# Analysis of missing data

We need to make sure that not only is the data equivalent between the intervention
groups but also that missing data is not unevenly distributed or excessive.
Reporting on the frequency of missing data stratified by intervention is also
easy, and to demonstrate this we make a data set with 10% of the placebo
arm having missing values, but 25% of the treatment arm: 

```{r}

# generate a dataset with values missing not at random compared to the intervention:
cgd_treat = cgd %>% mutate(treat = as.character(treat)) %>% filter(treat != "placebo")
cgd_placebo = cgd %>% mutate(treat = as.character(treat)) %>% filter(treat == "placebo")

set.seed(100)
mnar_cgd = bind_rows(
  cgd_placebo %>% .make_missing(p_missing = 0.1),
  cgd_treat %>% .make_missing(p_missing = 0.25)
)

```

Comparing this new data set we see that there is significant differences in some
of the data (but not the `steroids` variable). As this is quite a small dataset
it is not sufficiently powered to reliably detect the difference in missingness
at this level (15% difference).

```{r}
# compare the MNAR dataset against the intervention:
formula = Surv(tstart, tstop, status) ~ treat + 
  sex + age + height + weight + inherit + steroids + hos.cat

mnar_cgd %>% compare_missing(formula)

```

# Conversion of discrete data

Using this new data set with missing data it may be necessary to discretise some
or all of the data, or convert logical values into properly named factors.

```{r}

decade = function(x) sprintf("%d-%d",x-(x%%10),x-(x%%10)+9)

discrete_cgd = mnar_cgd %>% 
  # pick out the first episode
  filter(enum == 1) %>%
  # convert data
  make_factors(
    steroids,propylac,age,weight,height,
    .logical = c("recieved","not recieved"),
    .numeric = list(
      age="{decade(value)}",
      weight="{ifelse(value<20,'<20','20+')}",
      height="{ifelse(value<mean(value, na.rm=TRUE),'below average','above average')}"
    )
  )

formula = Surv(tstart, tstop, status) ~ treat + 
  sex + age + height + weight + inherit + steroids + hos.cat

old = options("tableone.show_pvalue_method"=TRUE)

# This comparison implicitly ignores missing values.
discrete_cgd %>% compare_population(formula)

options(old)
```

```R
# N.B. The following option is involved when converting integer data:
options("tableone.max_discrete_levels"=0)
# and is described in the documentation for make_factors().
```

# Making missing factors explicit:

In the comparison above missing values were not included, and we should be
cautious of the findings. Because of the missingness `tableone` will not 
calculate p-values. If factor values are missing (as in this case) then we
can include them as a new group and get a more robust comparison which includes
the distribution of missingness, and for which we can calculate a p-value. 

```{r}
discrete_cgd %>% explicit_na() %>% compare_population(formula)
```

# Non biomedical data

Beyond the bio-medical example `tableone` can make any more general comparison
between data that has a structure like:

`~ group + observation_1 + observation_2 + ... + observation_n`

We will use the `iris` and the `diamonds` datasets to demonstrate this more
general use case for `tableone`.

```{r}

# revert the labeller setting to the default
# and additionally hide the footer.
options("tableone.labeller"=NULL)
options("tableone.show_pvalue_method"=FALSE)
options("tableone.hide_footer"=TRUE)

# the heuristics detect that Petals in the iris data set are not normally
# distributed and hence report median and IQR:
iris %>% group_by(Species) %>% compare_population(tidyselect::everything())
```

The `missing_diamonds` data set which is included in this package has 10% of 
the values removed. This demonstrates the need for reporting the denominator.

```{r}
# The counts sometimes seem redundant if there is no missing information:
# however in a data set with missing values the denominators are important:
missing_diamonds %>% describe_population(tidyselect::everything())
```

